\MakeUppercase{Notes on Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-04 k-planar.tex
More correct explanations

Let $S_x$ (the separator at $x$) denote the set of all vertices $v\in V(G)$ such that $v$ appears at $x$ but does not appear at an ancestor of $x$.  We claim that $(S_x:x\in V(K))$ is a partition of $V(G)$. Indeed, if $v$ appears at nodes $x$ and $y$ where neither $x$ nor $y$ is an ancestor of the other, the $v$ contains a vertex of $Y_a$ for some common ancestor $a$ of $x$ and $y$.

@@

Next we describe a tree decomposition $(B_x:x\in V(K))$ of $H$.  (Don't be put off here by the fact that $V(H)=V(K)$).  We do this by definining, for each $x\in V(H)$, the subtree $K[x]=K[\{y:x\in B_y\}]$.  The subtree $K[x]$ contains $x$.  Additionally, for every edge $xz\in E(H)$, $K[x]$ includes the path from $x$ to $z$ in $K$.  This clearly produces a tree decomposition of $H$, what remains is to upper bound the width of this decomposition.

We claim that $K[x]$ is rooted at $x$, i.e., $x$ is the lowest common ancestor of all nodes in $B_x$.  To see why this is so, suppose $K[x]$ has root $a\neq x$.  In short: This happens because some $v\in V(N_x)$ is incident on an edge $vw\in E(G)$ that crosses $Y_a$.  Longer: This implies that $a$ has two children $b\neq c$ where $x$ is a descendant of $b$, some node $y$ is a descendant of $c$, and there is an edge $vw\in E(G)$ such that $v\in V(N_x)\setminus V(F_x)\subseteq V(N_b)\setminus V(F_a)$
and $w\in S_y\subset V(N_y)\subseteq V(N_b)\setminus V(F_a)$.  (Note, we allow $x=b$ and/or $y=c$ here.)  Therefore, $v$ and $w$ are in the interior of two different faces of the Mercedes graph $M_a$.  Therefore the edge $vw$ intersects $M_a$.  If it intersects $M_a$ at some vertex in $Y_a$, then $v$ would be in $S_a$, not $S_x$, a contradiction.  If it intersects $M_a$ at some vertex in $F_a$, then $v$ would be in $S_{a'}$ for some ancestor $a'$ of $a$, also a contradiction.

Now, fix some node $x\in V(K)$ and consider the contents of $B_x$.  As we have just argued, $B_x$ contains only ancestors of $x$ (including $x$ itself). If $B_x$ contains some ancestor $a$ of $x$, it is because there is an edge $vw\in E(G)$ with $v\in V(N_x)\setminus V(F(x))$ and $w\in S_a$.  Now, $w\in S_a$ because $w\in Y_a$ or there is an edge $wz\in E(G)$ that contains a point of $Y_a$.  This implies that there is a curve $C$ from $v$ to $z$ that contains at most $2k$ dummy  vertices of $G^+$.

>>>>

Let $S_x$ (the separator at $x$) denote the set of all vertices $v\in V(G)$ such that $v$ appears at $x$ but does not appear at an ancestor of $x$.  We claim that $(S_x:x\in V(K))$ is a partition of $V(G)$. Indeed, if $v$ appears at nodes $x$ and $y$ where neither $x$ nor $y$ is an ancestor of the other, then there exists edges $vw,vz\in E(G)$ (possibly $w=z$) such that $vw$ contains a vertex of $Y_x$ (as an endpoint or in its interior) and $vz$ contains a vertex of $Y_y$ (as endpoint or in its interior).  At least one of these two edges contains a vertex $Y_a$ for some common ancestor $a$ of $x$ and $y$.

@@

Next we describe a tree decomposition $(B_x:x\in V(K))$ of $H$.  (Don't be put off here by the fact that $V(H)=V(K)$).  We claim that, for every edge $xz\in V(H)$, $x$ is an ancestor of $z$ (or vice-versa).  Indeed, this follows from the same argument used to show that $S_x$ is a partition.  If neither $x$ nor $z$ is an ancestor of the other, then any edge $vw$ with $v\in S_x$ and $w\in S_z$ contains a vertex of $Y_a$ for some common ancestor $a$ of $x$ and $z$.

To describe our tree decomposition we define, for each $x\in V(H)$, the subtree $K[x]=K[\{y:x\in B_y\}]$.  The subtree $K[x]$ contains $x$.  Additionally, for every edge $xz\in E(H)$, where $x$ is an ancestor of $z$, $K[x]$ includes the path from $x$ to $z$ in $K$.  That this produces a tree-decomposition is due to the fact that for every edge $xz$ of $H$, $x$ is an ancestor of $z$ (or vice-versa).  What remains is to upper bound the width of this decomposition.

First observe that the subtree $K[x]$ is rooted at $x$, i.e., $x$ is the lowest common ancestor of all nodes in $B_x$.  Now, fix some node $x\in V(K)$ and consider the contents of $B_x$.  As we have just argued, $B_x$ contains only ancestors of $x$ (including $x$ itself). If $B_x$ contains some ancestor $a$ of $x$, it is because there is an edge $vw\in E(G)$ with $v\in V(N_x)\setminus V(F(x))$ and $w\in S_a$.  Now, $w\in S_a$ because $w\in Y_a$ or there is an edge $wz\in E(G)$ that contains a point of $Y_a$.  This implies that there is a curve $C$ from $v$ to $z$ that contains at most $2k$ dummy  vertices of $G^+$.
